---
title: "HW2"
author: "Valeria"
date: "2024-05-14"
output:
  html_document:
    keep_md: yes
  pdf_document: default
---

Libraries
```{r}
library(tidyr)
library(dplyr)
library(vegan)
library(ggplot2)
library(gridExtra)
library(ape)
library(pvclust)
library(golubEsets)
```

## data

The data are from Golub et al. These are the combined training samples and test samples. There are 47 patients with acute lymphoblastic leukemia (ALL) and 25 patients with acute myeloid leukemia (AML).

```{r}
golub <- data.frame(Golub_Merge)[1:7129]
rownames(golub) <- paste(Golub_Merge$ALL.AML, rownames(golub), sep = "_")

#group vector to check
Golub_Merge$ALL.AML

```

```{r}
summary(golub)
```
```{r}
if(sum(is.na(golub)) > 0) {
  cat("There are missing values in the dataset")
} else {
  cat("No missing values in the dataset")
}
```
## Density plot
```{r}

golub %>%
  colMeans() %>%
  as_tibble() %>%
  arrange(value) %>%
  ggplot(aes(x = value)) +
  geom_density() +
  xlab("Raw data means")

```
The graph is not exactly symmetrical, it is slightly shifted to the right
##normalization


```{r}
log_golub <-  decostand(golub, method = "log", MARGIN = 2)
log_golub %>%
  colMeans() %>%
  as_tibble() %>%
  arrange(value) %>%
  ggplot(aes(x = value)) +
  geom_density() +
  xlab("Log norm data means")


```
```{r}
stnd_golub <-  decostand(golub, method = "standardize", MARGIN = 2)
stnd_golub %>%
  colMeans() %>%
  as_tibble() %>%
  arrange(value) %>%
  ggplot(aes(x = value)) +
  geom_density() +
  xlab("Standatrization data means")


```
Log normilized data looks better 
##Distances
```{r}
euclidean_dist <- vegdist(log_golub, method = "euclidean")
manhattan_dist <- vegdist(log_golub, method = "manhattan")
canberra_dist <- vegdist(log_golub, method = "canberra")
gower_dist <- vegdist(log_golub, method = "gower")

# Set up the plotting layout for better visualization
par(mfrow = c(2, 3))

# Plot histograms of the distances with customized colors and labels
hist(euclidean_dist, main = "Euclidean Distance", xlab = "Distance", col = "skyblue", border = "black")
hist(manhattan_dist, main = "Manhattan Distance", xlab = "Distance", col = "lightgreen", border = "black")
hist(canberra_dist, main = "Canberra Distance", xlab = "Distance", col = "salmon", border = "black")
hist(gower_dist, main = "Gower Distance", xlab = "Distance", col = "orange", border = "black")

# Add a common title for all plots
mtext("Comparison of Different Distance Metrics", line = 0.5, cex = 1.2, outer = TRUE)
```


```{r}
disatances <- data.frame(Euclidean = as.numeric(euclidean_dist), Manhattan = as.numeric(manhattan_dist), Canberra = as.numeric(canberra_dist),Gower = as.numeric(gower_dist))
Pl_hist <- ggplot(disatances) + geom_density()

Pl_euclidean <- Pl_hist + aes(x = Euclidean) + labs(x = "euclidean")
Pl_manhattan <- Pl_hist + aes(x = Manhattan) + labs(x = "manhattan")
Pl_gower <- Pl_hist + aes(x = Gower) + labs(x = "gower")
Pl_canberra <- Pl_hist + aes(x = Canberra) + labs(x = "canberra")
grid.arrange(Pl_euclidean, Pl_manhattan, Pl_gower, Pl_canberra)
```

##Clustering
```{r}

clustering <- function(dist) {
  dists <- c("Euclidean", "Manhattan", "Canberra", "Gower")
  methods <- c("single", "complete", "average", "ward.D2")
  
  for (m in methods) {
    hclust <- hclust(dist, method = m)
    corr <- cor(dist, as.dist(cophenetic(hclust)))
    cat("nClustering Method:", m, "n")
    print(corr)
    }
}

# Calculating and plotting clustering for different distance metrics
clustering(euclidean_dist)
clustering(manhattan_dist)
clustering(canberra_dist)
clustering(gower_dist)

# Performing hierarchical clustering using Euclidean distance
avg_euc <- hclust(euc, method = "average")
ph_avg_euc <- as.phylo(avg_euc)
plot(ph_avg_euc, type = "phylogram", main = "Hierarchical Clustering using Euclidean Distance")
axisPhylo()

# Performing hierarchical clustering using Manhattan distance
single_man <- hclust(manhattan_dist, method = "single")
ph_man <- as.phylo(single_man)
plot(ph_man, type = "phylogram", main = "Hierarchical Clustering using Manhattan Distance")
axisPhylo()

# Performing hierarchical clustering using Canberra distance
hclust_canb <- hclust(canb, method = "complete")
ph_canb <- as.phylo(hclust_canb)
plot(ph_canb, type = "phylogram", main = "Hierarchical Clustering using Canberra Distance")
axisPhylo()

# Performing hierarchical clustering using Gower distance

hclust_gower <- hclust(gower_dist, method = "average") # Using "average" method for Gower distance
ph_gower <- as.phylo(hclust_gower)
plot(ph_gower, type = "phylogram", main = "Hierarchical Clustering using Gower Distance")
axisPhylo()
```
Clustering using Manhattan distance looks the most strange, Euclidian and Gower metrics have some common.
```{R}
# Defining methods and distances
distances <- list(Euclidean = euclidean_dist, Manhattan = manhattan_dist, Canberra = canberra_dist, Gower = gower_dist)

methods <- c("single", "complete", "average", "ward.D2")

# Function to perform clustering and evaluate its quality
clustering_evaluation <- function(data, indices, clustering_method) {
  # Performing clustering
  clusters <- clustering_method(data)
  
  # Evaluating clustering quality using correlation coefficient
  correlation <- cor(as.numeric(group_vector[indices]), clusters$order, method = "spearman")
  
  return(correlation)
}

# Function to compare clustering methods using correlation test
correlation_test <- function(distances, methods) {
  results <- data.frame(distance_metric = character(), clustering_method = character(), correlation_value = numeric())
  
  for (dist_name in names(distances)) {
    dist <- distances[[dist_name]]
    for (method in methods) {
      cat("Distance:", dist_name, "Method:", method, "\n")
      correlation_value <- clustering_correlation_test(dist, method)
      results <- rbind(results, data.frame(distance_metric = dist_name, clustering_method = method, correlation_value = correlation_value))
    }
  }
  
  # Identifying which metric is better based on the highest correlation value
  best_metric <- results[which.max(results$correlation_value), "distance_metric"]
  cat("The best distance metric is:", best_metric)
}

# Function to perform clustering and correlation test
clustering_correlation_test <- function(dist, method) {
  hclust_result <- hclust(dist, method = method)
  corr <- cor(dist, as.dist(cophenetic(hclust_result)))
  cat("Clustering Correlation:", corr, "\n")
  return(corr)
}

# Call the correlation test function with the defined distances and methods
correlation_test(distances, methods)

```
The best distance metric is Manhattan, althow the plot was the most strange. 

```{r, fig.width=10, fig.height=5}
plot(pvclust(t(golub), method.hclust = "average", nboot = 70, 
             method.dist = "manhattan", parallel = TRUE, iseed = 100))


```

Сlustering analysis on samples of Acute Myeloid Leukemia (AML) and Acute Lymphoblastic Leukemia (ALL) was performed to investigate the separation of the two types of leukemia based on gene expression data.

Results:
After clustering the samples and according to the bootstrap results, almost all the samples of AML were successfully separated from ALL based on gene expression patterns. However, it was noted that some samples still appeared mixed up and showed overlap between the AML and ALL clusters.

If we consider that there were no mistakes in the samles and no contamination, it can be possible, that these types of cansers have some genes with the same expression profile [1]. Also, there could be a genetic heterogeneity within the samples. Maybe the problem was in preprocessing or batch effect could happen.

##References. 
1. https://www.sciencedirect.com/science/article/abs/pii/S0344033815001491